Convergent Adaptation while adapting

### 2014-07-02 patchy-selection-paper.tex
addition to applications

The convergent evolution of similar phenotypes in response to shared
selection pressures is a testiment to the power of selection to
repeatedly sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level, 
with disparate taxa converging to the same phenotype through parallel 
genetic changes in the same pathway, genes, 
or even by precisely the same genetic changes \citep{zhen2012parallel}. 
Such precise convergence points to the conservation of function of the genes underlying adaptive traits, 
sometimes over deep time scales \citep{zhen2012parallel,martin2013repeated,deephomologypapers}. 
Convergence on a molecular level also suggests that the path 
accessible to adaptation is sometimes relatively constrained \citep{stern2013genetic}, 
as would be the case if only a few changes confer the advantagous phenotype 
and are sufficiently free of deleterious pleiotropic consequences.

%For highly polygenic traits

Convergent adaptation can also occur within species, if individuals within a species 
adapt to the same environment, in parallel, through different genotypic changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes REFs.
Such evolution of convergent phenotypes is favored by higher mutational input, 
i.e.\ larger mutational target and/or population size \citep{softsweepsII}.
The geographic distribution of populations can also affect the probability of parallel mutation within a species:
a widespread species is more likely to adapt by multiple, parallel mutations if dispersal is geographically limited, 
since subpopulations will adapt via new mutations before the adaptive allele arrives via migration \citep{ralph2010parallel}. 
Standing variation for a trait can also increase the probability of convergence, 
and standing variation for an allele increases the probability that the selective sweep will be \emph{soft}
(beginning from a base of multiple copies),
leading to genetic patterns similar to convergent adaptation \citep{orr2001sieve,softsweepsI}.

@@

In this paper, we study this situation, namely,
when a set of alleles provide an adaptive benefit in geographically localized patches, 
separated by (inhabited) regions where the alleles are deleterious.
The main questions are:
Under what conditions is it likely that each patch adapts in parallel, through new mutation,
and when is it likely that migration carries these alleles between patches?
How could we tell, after the fact, which has occurred?
We work in a model of continuous geography,
using a variety of known results and new methods.
In section XX we \dots QUICK OUTLINE

@@

%%%%%%%%%%%%
\subsection{Multiple patches}
Above we worked with two patches separated by distance $R$;
but these calculations apply as well to multiple patches
in which the $B$ allele can facilitate local adaptation. 
Under our assumptions of mutational exclusion and rapid establishment, 
each patch will adapt through establishment of a single copy of the $B$ allele, 
either by migration or mutation.
If we imagine that the times between successive establishments is the result of many nearly-independent attempts
with small probability of success,
the process of occupation is well approximated by a Markov chain,
whose state records the labeling of patches according to the colonizing allele
(or none, for those that haven't yet adapted).

If not yet adapted, a patch of area $A$ will acquire the adaptation by new mutation at rate $2 s_p \mu \rho A/\xi^2$ (equation \eqref{eqn:mutrate}).
Suppose that patches $1, \ldots, k$ are already adapted,
and that these patches are at distances $r_1, \ldots, r_k$ away from this unadapted patch.
The total rate of adaptation through migration, from equation \eqref{eqn:migrate}, is
\begin{equation}
  \migrate(r_i) = C \frac{ \rho s_p }{\xi^2 s_m} \sum_{i=1}^{k} \left(r_i \sqrt{2 s_m} /\sigma \right)^{-(d-1)/2} \exp\left(- r_i \sqrt{2 s_m} /\sigma\right),
\end{equation}
and if the patch adapts through migration, the probability the adapted allele 
comes from patch $i$ is 
\begin{align}  \label{eqn:prob-parallel}
  \frac{\left(r_i \right)^{-(d-1)/2} \exp\left(- r_i \sqrt{2 s_m}
  /\sigma\right)} {\sum_{j=1}^{k}  \left(r_i \right)^{-(d-1)/2} \exp\left(- r_i \sqrt{2 s_m}
    /\sigma\right) } .
\end{align}
These specify the transition rates of the process.

@@

Above we argued that a good model for this transit is a single Brownian trunk lineage
surrounded by a cloud of close relatives of random size $K$
whose chance of dying out by $t$ generations is $k_e(t)$.
Consider a single such family, and let $\tau$ be the (random) time at which it first hits the new patch,
with $\tau = \infty$ if it never does. 
We assume that the length of the hitchhiking segment depends only on 
the time spent by the trunk lineage of the first successful migrant ``in transit'' between the patches,
i.e.\ $\tau$ conditioned on $\tau < \infty$.
In so doing, we neglect 
recombination between $B$ alleles (since they are at low density in transit, we assume this is rare),
and the possibility that more than one migrant family is in transit at once
(so that faster migrants would be more likely to have arrived first);
but these factors should be minor if the rest of our framework fits.

@@

Based on this picture of
the lineage bouncing around in the patch,
we propose a simple heurstic to describe the probability of recombination. 
If recombination is on a slower time scale than mixing within the patch,
then the time until the last recombination is the integral of $r (1-p(x)) (1+s(x)/2)$ 
against the stationary distribution (equation~\ref{eqn:lineagestatdist}).
Within the patch, $p(x) \approx 1$,
so recombination off the $B$ background is unlikely to occur,
and so the rate of recombination depends to a first approximation 
on the proportion of the patch in which $p(x)$ is intermediate,
i.e.\ where both $B$ and $b$ are present.
This region where $p(x)$ is intermediate has width of order $\sigma/\sqrt{s}$, where $s = \min(s_p,s_m)$.
As we increase the size of a patch, the area on which $p(x)$ is intermediate scales with the perimeter of the patch.
Therefore, the proportion of the time a lineage at stationarity is found in the intermediate regions
multiplied by $p(x)$ in these intermediate regions, where we assume
$p(x) \approx 1/2$. Under these approximations the per generation rate of
recombination of a linked lineage to the $b$ background is
reduced by a factor roughly equal to the proportion of the patch lying within distance $\sigma/\sqrt{s}$ of the boundary,
so the effective recombination rate is
% \plr{Note to ourselves: If a $d$-dimensional object with area A has radius $r$ (in any sense)
% and you shrink it to have radius $r_0$,
% then the resulting object has area $A \times (r_0/r)^d$;
% here we have $r_0 = r - \sigma/\sqrt{s}$,
% in $d=1$ the radius is $A/2$ and in $d=2$ it is $\sqrt{A/\pi}$.}
\begin{align}
r_\text{eff} &=
    \begin{cases}
    r \frac{ \sigma /\sqrt{s} }{ \sigma/\sqrt{s} + A }    \qquad & d=1 \\
   r \frac{\pi  \sigma /\sqrt{s}  \left( \sigma /\sqrt{s}+
  \sqrt{A/\pi} \right)}{A+\pi \sigma/\sqrt{s} \left( \sigma/\sqrt{s} +
  \sqrt{A/\pi} \right) }    \qquad & d=2 \\  
\end{cases}
\label{eqn:rate-rec-in-patch}   %(\sigma/\sqrt{s})A^{-1/d}
\end{align}
which scales the recombination rate roughly with $(\sigma/\sqrt{s})A^{-1/d}$.
We can use this to think about the length of the haplotype
associated with the $B$ allele after a given length of time $T$,
for instance, the time since the $B$ allele joined the patch.
% If the recombination rate per base is $r_{BP}$, 
% then the length of the haplotype associated with the $B$ allele is
% \begin{equation}
% \frac{\sqrt{s}}{\sigma A^{-1/d}}\frac{1}{r_{BP}}
% \end{equation} 
In units of genetic length (Morgans),
the length of the haplotype associated with the $B$ allele is, up to a constant,
\begin{equation}
  \frac{A^{1/d}\sqrt{s}}{\sigma T}
\end{equation} 
Note that the length of haplotype decays at a rate proportional to the
cline width $\sqrt{s} / \sigma$, as the narrower the cline the less
opportunity for recombination to move our lineage off the $B$ background.

Finally we can consider the expected length of haplotype shared within a patch. 
To do this, we compare the rate of recombination between $B$ and $b$ haplotypes (equation~\ref{eqn:rate-rec-in-patch})
to the rate of coalescence of $B$ haplotypes within a patch. 
We will assume that mixing within the patch is fast compared to the rate of drift, 
so that the rate of coalescence of $B$ alleles within a patch is 
$\xi^2/(A\rho)$,
the ``variance effective population size'' of the patch
(this bypasses the well-known difficulties to spatial coalescence \citep{felsenstein1975pain,barton2002neutral}).
The rate of recombination and coalescence will be on the same order if
$\xi^2 /(A\rho) \approx r (\sigma/\sqrt{s}) A^{-1/d}$, 
and hence the genetic length of haplotype shared between pairs of
individuals within the patches is on the order of
\begin{align}
  r = \frac{\sqrt{s} \xi^2}{\sigma \rho A^{(d-1)/d}} 
\end{align}
which does not depend on $A$ in one dimension,
but decreases with $\sqrt{A}$ in two dimensions.
In applications, this can be compared to equation~\eqref{eqn:haplotype_length}
for the length of shared haplotype between patches.

%%%%%%%%%%
\section{Applications} 
Hoekstra, Nachman, and colleagues have made an extensive study of coat color in the rock pocket mouse, 
\emph{Chaetodipus intermedius}, building on classic work by \citet{benson1933concealing} and \citet{dice1940ecologic}.
These mice live on rocky outcrops throughout the southwestern United States and northern Mexico, 
and generally have a light pelage similar to the predominant rock color.
However, in some regions these mice live on darker substrates (e.g.\ old lava flows),
and in these patches have much darker pigmentation, 
presumably to avoid visual predators.
\citet{nachman2003genetic} demonstrated that on one of these flows (Pinacate), 
much of the change in coloration is due to an allele at the MC1R locus.
This dark allele differs from the light allele by four amino acid changes, 
and has a dominant or partially dominant effect depending on the measure of coat color assessed. 
These areas of dark rock are separated from each other by a range of distances, 
with some being separated from any other by hundreds of kilometers of light colored substrate. 
The Pinacate allele is not present in a number of other populations of dark colored rock pocket mouse, 
suggesting these populations have adapted in parallel \citep{hoekstra2003different}.
However, there is some evidence \citep{Hoekstra:05} that, elsewhere in the range, 
multiple dark outcrops may share a dark phenotype whose genetic basis has been spread by migration
despite intervening light habitat. 
We apply our theory to this example,
in particular investigating how the probability of parallel mutation depends on 
the distance between lava flow patches.

A key parameter above was the dispersal distance divided by the square root of strength of selection against the focal allele between patches, $\sigma/\sqrt{s_m}$. 
\citep{hoekstra2004ecological} studied the frequency of the dark MC1R allele and coat color phenotypes, 
at sites across the (dark) Pinacate lava flow and at two nearby sites in light-colored rock.
On the lava flow the dark MC1R allele is at 86\% frequency,
while at a site with light substrate 12km to the west (Tule), the frequency is 3\%.
The dark allele was entirely absent from Christmas pass, a site with light substrate 7km north of Tule, and 3km further from the lava flow.
In the other direction, the dark MC1R allele was at 34\% at a site with light substrate 10km to the east of the flow (O'Neill).
These numbers give us a sense of a plausible range of characteristic lengths.
Assuming the dark allele is at 50\% frequency at the edge of the lava flow, 
we can fit formula \eqref{eqn:eqfreq} to these frequencies
\citep[similar to][]{hoekstra2004ecological}.
Doing this for Tule we obtain $\sigma/\sqrt{s_m} \approx 3$km, 
and for O'Neill, $\sigma/\sqrt{s_m} \approx 30$km, giving us a range of cline widths. 
We also need estimate of $s_m$. 
Using $\sigma=1km$ \gc{REF}, these cline widths imply $s_m=1/\sqrt{3}$ and
$1/\sqrt{30}$ are reasonable values.
\plr{Going back to previous conversations, we came upon \citet{mullen2008natural}, 
which has dispersal distance of 0.26km.  Shall we ask Nachman?}
\plr{He referred to \citet{french1968dispersal,allred1963range}.}

The mutational target size $\mu$ for the trait is unclear. 
While the Pinacate dark haplotype differs from the light haplotype at three amino-acid residues,
it is likely that not all of these changes are needed for a population to begin to adapt. 
Also there a number of genes besides MC1R at which adaptative changes affecting pigmentation 
have been identified in closely related species and more broadly across vertebrates \citep{REF}.
To span a range of plausible values, we use a low mutation rate of $\mu= 10^{-8}$, 
where the adaptive mutation has to occur at a single base pair; 
and a high mutation rate $\mu= 10^{-5}$, equivalent to the mutation
being able to arise at any of a thousand base pairs.  
Finally, we set $A=100\text{km}^2$.
Figure~\ref{fig:mice_prob_parallel} shows the dependance of the probability
of parallel mutation on the distance between lava flow patches using these parameters, 
showing that parallel mutation should become
likely over a scale of tens to a few hundred kilometers between patches.

\begin{figure}[ht]
  \begin{center}
    \includegraphics{Lava_flow_mice_prob_parallel}
  \end{center}
  \caption{
The probability of parallel mutation as a function of the distance
between environmental patches for two different cline widths
and two different mutation rates (using equation \eqref{eqn:prob-parallel}). 
The parameters were chosen to match  XXX
  \label{fig:mice_prob_parallel}
  }
\end{figure}

Givn the large selection coefficient associated with the dark allele
on the dark substrate we expect the initial sweep associated with
either a new mutation or migrant allele to be large ($s/\log(\rho A
s)$). We can use eqn \eqref{eqn:haplotype_length} to calculate how long a
founding haplotype we initially expect to find shared between populations
this in Figure \ref{fig:mice_prob_parallel}. The initial length can be
quite long between geographically close patches (tens of km). However,
for our wider cline width ($\sigma/\sqrt{s_m} =30km$) adaptation by
migration can still be likely for patches 100km apart but its shared
basis may be hard to detect as the length of shared haplotype can be
quite short.

Given the relatively small size of the dark substrate patches where
$A$ is often $<10km$ we should expect the haplotype to be rapidly
whittled down within patches (eqn. \ref{})

@@

% patchy selection helps this happen; 
% effective migration rate looks like what; 
% strength of (positive?) selection suprisingly does not affect results (?)

% standing variation will be important under these circumstances; 
% migration rate (and so geographic structure) will not affect numbers of types if this is true; 
% however migration always affcts spatial patterns.

% recessive alleles will have different dynamics. Actually not really
% obvious what will happen here, cline width will be much wider but
% prob. of taking off will go as the square of the occupancy density.

@@

% small-scale fluctuations
% when landscape is not made of discrete (circular) patches

>>>>

% Such precise convergence points to the conservation of function of the genes underlying adaptive traits, 
% sometimes over deep time scales \citep{deephomologypapers}. 
% Convergence on a molecular level also suggests that the path 
% accessible to adaptation is sometimes relatively constrained \citep{stern2013genetic}, 
% as would be the case if only a few changes confer the advantagous phenotype 
% and are sufficiently free of deleterious pleiotropic consequences.

The convergent evolution of similar phenotypes in response to shared
selection pressures is a testiment to the power of selection to
repeatedly sculpt phenotypic variation. 
In some cases this convergence extends to the molecular level, 
with disparate taxa converging to the same phenotype through parallel 
genetic changes in the same pathway, genes, 
or even by precisely the same genetic changes \citep{zhen2012parallel,,martin2013repeated,deephomologypapers}. 
Convergent adaptation can also occur within species, if individuals within a species 
adapt to the same environment, in parallel, through different genetics changes. 
There are a growing number of examples of this in a range of well studied organisms and phenotypes REFs.
Such evolution of convergent phenotypes is favored by higher mutational input, 
i.e.\ larger mutational target and/or population size \citep{softsweepsII}.
The geographic distribution of populations can also affect the probability of parallel mutation within a species:
a widespread species is more likely to adapt by multiple, parallel mutations if dispersal is geographically limited, 
since subpopulations will adapt via new mutations before the adaptive allele arrives via migration \citep{ralph2010parallel}. 
Standing variation for a trait can also increase the probability of convergence, 
and standing variation for an allele increases the probability that the selective sweep will be \emph{soft}
(beginning from a base of multiple copies),
leading to genetic patterns similar to convergent adaptation \citep{orr2001sieve,softsweepsI}.

@@

In this paper, we study this situation, namely,
when a set of alleles provide an adaptive benefit in geographically localized patches, 
separated by (inhabited) regions where the alleles are deleterious.
The main questions are:
Under what conditions is it likely that each patch adapts in parallel, through new mutation,
and when is it likely that migration carries these alleles between patches?

How could we tell, after the fact, which has occurred?

We work in a model of continuous geography,
using a variety of known results and new methods.
In section \ref{ss:patchymutation} we develop a simple approximation,
eqn. \eqref{eqn:mutrate}, for the rate at which patches become
established by new mutations. Then in section \ref{ss:patchymigration} we
develop an approximation, eqn. \eqref{eqn:migrate}, to the rate which alleles move from an
existing patch to colonize a novel patch (this represents a big chunk
of the results of the paper). We combine these two results in section
\ref{ss:probparallel} to develop an approximation to the probability of
parallel adaptation, eqn. \eqref{}.
To understand the genomic signal left by adaptations shared by
migration in section \ref{ss:haplotype_length} we approximate the time
it will take for an allele to transit between patches,
eqn. \eqref{eqn:mean_tau}, and use this to approximate the length of
haplotype that hitchhikes with it,
eqn. \eqref{eqn:haplotype_length}. We then quantify how quickly this
haplotype decays within patches eqn. \eqref{} in section \ref{ss:patch_haplotype}.

\dots QUICK OUTLINE

@@

%%%%%%%%%%%%
\subsubsection{Multiple patches}
Above we worked with two patches separated by distance $R$;
but these calculations apply as well to multiple patches
in which the $B$ allele can facilitate local adaptation. 
Under our assumptions of mutational exclusion and rapid establishment, 
each patch will adapt through establishment of a single copy of the $B$ allele, 
either by migration or mutation.
If we imagine that the times between successive establishments is the result of many nearly-independent attempts
with small probability of success,
the process of occupation is well approximated by a Markov chain,
whose state records the labeling of patches according to the colonizing allele
(or none, for those that haven't yet adapted).

If not yet adapted, a patch of area $A$ will acquire the adaptation by new mutation at rate $2 s_p \mu \rho A/\xi^2$ (equation \eqref{eqn:mutrate}).
Suppose that patches $1, \ldots, k$ are already adapted,
and that these patches are at distances $r_1, \ldots, r_k$ away from this unadapted patch.
The total rate of adaptation through migration, from equation \eqref{eqn:migrate}, is
\begin{equation}
  \migrate(r_i) = C \frac{ \rho s_p }{\xi^2 s_m} \sum_{i=1}^{k} \left(r_i \sqrt{2 s_m} /\sigma \right)^{-(d-1)/2} \exp\left(- r_i \sqrt{2 s_m} /\sigma\right),
\end{equation}
and if the patch adapts through migration, the probability the adapted allele 
comes from patch $i$ is 
\begin{align}  
  \frac{\left(r_i \right)^{-(d-1)/2} \exp\left(- r_i \sqrt{2 s_m}
  /\sigma\right)} {\sum_{j=1}^{k}  \left(r_i \right)^{-(d-1)/2} \exp\left(- r_i \sqrt{2 s_m}
    /\sigma\right) } .
\end{align}
These specify the transition rates of the process.

@@

Above we argued that a good model for this transit is a single Brownian trunk lineage
surrounded by a cloud of close relatives of random size $K$
whose chance of surviving until $t$ generations is $1-k_e(t)$.
Consider a single such family, and let $\tau$ be the (random) time at which it first hits the new patch,
with $\tau = \infty$ if it never does. 
We assume that the length of the hitchhiking segment depends only on 
the time spent by the trunk lineage of the first successful migrant ``in transit'' between the patches,
i.e.\ $\tau$ conditioned on $\tau < \infty$.
In so doing, we neglect 
recombination between $B$ alleles (since they are at low density in transit, we assume this is rare),
and the possibility that more than one migrant family is in transit at once
(so that faster migrants would be more likely to have arrived first);
but these factors should be minor if the rest of our framework fits.

@@

Based on this picture of
the lineage bouncing around in the patch,
we propose a simple heurstic to describe the probability of recombination. 
If recombination is on a slower time scale than mixing within the patch,
then the time until the last recombination is the integral of $r (1-p(x)) (1+s(x)/2)$ 
against the stationary distribution (equation~\ref{eqn:lineagestatdist}).
Within the patch, $p(x) \approx 1$,
so recombination off the $B$ background is unlikely to occur,
and so the rate of recombination depends to a first approximation 
on the proportion of the patch in which $p(x)$ is intermediate,
i.e.\ where both $B$ and $b$ are present.
This region where $p(x)$ is intermediate has width of order $\sigma/\sqrt{s}$, where $s = \min(s_p,s_m)$.
As we increase the size of a patch, the area on which $p(x)$ is intermediate scales with the perimeter of the patch.
Therefore, the proportion of the time a lineage at stationarity is found in the intermediate regions
multiplied by $p(x)$ in these intermediate regions, where we assume
$p(x) \approx 1/2$. Under these approximations the per generation rate of
recombination of a linked lineage to the $b$ background is
reduced by a factor roughly equal to the proportion of the patch lying within distance $\sigma/\sqrt{s}$ of the boundary,
so the effective recombination rate is
% \plr{Note to ourselves: If a $d$-dimensional object with area A has radius $r$ (in any sense)
% and you shrink it to have radius $r_0$,
% then the resulting object has area $A \times (r_0/r)^d$;
% here we have $r_0 = r - \sigma/\sqrt{s}$,
% in $d=1$ the radius is $A/2$ and in $d=2$ it is $\sqrt{A/\pi}$.}
\begin{align}
r_\text{eff} &=
    \begin{cases}
    r \frac{ \sigma /\sqrt{s} }{ \sigma/\sqrt{s} + A }    \qquad & d=1 \\
   r \frac{\pi  \sigma /\sqrt{s}  \left( \sigma /\sqrt{s}+
  \sqrt{A/\pi} \right)}{A+\pi \sigma/\sqrt{s} \left( \sigma/\sqrt{s} +
  \sqrt{A/\pi} \right) }    \qquad & d=2 ,  
\end{cases}
\label{eqn:rec_rate_within}   %(\sigma/\sqrt{s})A^{-1/d}
\end{align}
so that $T$ generations after the founding of a patch we expect a
haplotype of genetic length $1/(r_\text{eff}T)$ Morgans to remain linked to
the $B$ allele in a typical lineage. 
Note that the length of haplotype
decays at a rate roughly proportional to $\sqrt{s}/ \sigma$, as the narrower the cline the less
opportunity for recombination to move our lineage off the $B$ background.

Finally we can consider the expected length of haplotype shared within a patch. 
To do this, we compare the rate of recombination between $B$ and $b$ haplotypes (equation~\ref{eqn:rec_rate_within})
to the rate of coalescence of $B$ haplotypes within a patch. 
We will assume that mixing within the patch is fast compared to the rate of drift, 
so that the rate of coalescence of $B$ alleles within a patch is 
$\xi^2/(A\rho)$,
the ``variance effective population size'' of the patch
(this bypasses the well-known difficulties to spatial coalescence \citep{felsenstein1975pain,barton2002neutral}).
The rate of recombination and coalescence will be on the same order if
$\xi^2 /(A\rho) \approx r (\sigma/\sqrt{s}) A^{-1/d}$, 
and hence the genetic length of haplotype shared between pairs of
individuals within the patches at equilibrium is on the order of
\begin{align} \label{eqn:within_haplotype_length}
  r = \frac{\sqrt{s} \xi^2}{\sigma \rho A^{(d-1)/d}} 
\end{align}
which does not depend on $A$ in one dimension,
but decreases with $\sqrt{A}$ in two dimensions.
In applications, this can be compared to equation~\eqref{eqn:haplotype_length}
for the length of shared haplotype between patches.

%%%%%%%%%%
\section{Applications}

% FOR REFERENCE:
% On Wed, May 28, 2014 at 10:20 AM, Michael Nachman <mnachman@berkeley.edu> wrote:
% > Graham:
% >
% > The density is probably in the range of 10-40 animals per hectare (10,000
% > m^s).  We built enclosures of 0.25 hectare and found that they could support
% > about 10 animals with food provided ad libitum, so I think this is an upper
% > density (i.e. 40/hectare).  As for dispersal distance, I'm less sure.
% > Studies on closely related species (genus Perognathus) in similar habitats
% > suggest distances of 100's of feet.  Attached are some old papers that tried
% > to estimate dispersal distances in Perognathus.
% > best,
% > Michael

Hoekstra, Nachman, and colleagues have made an extensive study of coat color in the rock pocket mouse, 
\emph{Chaetodipus intermedius}, building on classic work by \citet{benson1933concealing} and \citet{dice1940ecologic}.
These mice live on rocky outcrops throughout the southwestern United States and northern Mexico, 
and generally have a light pelage similar to the predominant rock color.
However, in some regions these mice live on darker substrates (e.g.\ old lava flows),
and in these patches have much darker pigmentation, 
presumably to avoid visual predators.
\citet{nachman2003genetic} demonstrated that on one of these flows (Pinacate), 
much of the change in coloration is due to an allele at the MC1R locus.
This dark allele differs from the light allele by four amino acid changes, 
and has a dominant or partially dominant effect depending on the measure of coat color assessed. 
These areas of dark rock are separated from each other by a range of distances, 
with some being separated from any other by hundreds of kilometers of light colored substrate. 
The Pinacate allele is not present in a number of other populations of dark colored rock pocket mouse, 
suggesting these populations have adapted in parallel \citep{hoekstra2003different}.
However, there is some evidence \citep{Hoekstra:05} that, elsewhere in the range, 
multiple dark outcrops may share a dark phenotype whose genetic basis has been spread by migration
despite intervening light habitat. 
We apply our theory to this example,
in particular investigating how the probability of parallel mutation depends on 
the distance between lava flow patches.

A key parameter above was the dispersal distance divided by the square root of strength of selection against the focal allele between patches, $\sigma/\sqrt{s_m}$. 
\citet{hoekstra2004ecological} studied the frequency of the dark MC1R allele and coat color phenotypes, 
at sites across the (dark) Pinacate lava flow and at two nearby sites in light-colored rock.
On the lava flow the dark MC1R allele is at 86\% frequency,
while at a site with light substrate 12km to the west (Tule), the frequency is 3\%.
The dark allele was entirely absent from Christmas pass, a site with light substrate 7km north of Tule, and 3km further from the lava flow.
In the other direction, the dark MC1R allele was at 34\% at a site with light substrate 10km to the east of the flow (O'Neill).
These numbers give us a sense of a plausible range of characteristic lengths.
Assuming the dark allele is at 50\% frequency at the edge of the lava flow, 
we can fit formula \eqref{eqn:eqfreq} to these frequencies
\citep[similar to][]{hoekstra2004ecological}.
Doing this for Tule we obtain $\sigma/\sqrt{s_m} \approx 3$km, 
and for O'Neill, $\sigma/\sqrt{s_m} \approx 30$km, giving us a range of cline widths. 
We also need estimate of $s_m$. 
Using $\sigma \approx 1$km \citep{french1968dispersal,allred1963range}, 
these cline widths imply that $s_m=1/\sqrt{3}$ and $1/\sqrt{30}$ are reasonable values.

\begin{figure}[ht]
  \begin{center}
    \includegraphics{Lava_flow_mice_prob_parallel}
  \end{center}
  \caption{
The probability of parallel mutation as a function of the distance
between environmental patches for two different cline widths
and two different mutation rates (using equation \eqref{eqn:parallel_prob}). 
The parameters were chosen to match the example of \textit{Chaetodipus intermedius};
see text.
  \label{fig:mice_prob_parallel}
  }
\end{figure}

The mutational target size $\mu$ for the trait is unclear. 
While the Pinacate dark haplotype differs from the light haplotype at three amino acid residues,
it is likely that not all of these changes are needed for a population to begin to adapt. 
Also there a number of genes besides MC1R at which adaptative changes affecting pigmentation 
have been identified in closely related species and more broadly across vertebrates \citep{Hoekstra:06}.
To span a range of plausible values, we use a low mutation rate of $\mu= 10^{-8}$, 
where the adaptive mutation has to occur at a single base pair; 
and a high mutation rate $\mu= 10^{-5}$, equivalent to the mutation
being able to arise at any of a thousand base pairs.  
Finally, we set $A=100\text{km}^2$ (roughly the size of the Pinacate patch).
Figure~\ref{fig:mice_prob_parallel} shows the dependance of the probability
of parallel mutation on the distance between lava flow patches using these parameters, 
showing that parallel mutation should become
likely over a scale of tens to a few hundred kilometers between patches.

Given the large selection coefficient associated with the dark allele
on the dark substrate, we expect the initial haplotype associated with
either a new mutation or migrant allele to be large. 
We can use equation \eqref{eqn:haplotype_length} to calculate how long the founding
haplotype shared between populations is expected to be,
as shown in Figure~\ref{fig:mice_prob_parallel}. 
The initial length can be quite long between geographically close patches (tens of kilometers). 
However, for the wider cline width ($\sigma/\sqrt{s_m} = 30$km), 
adaptation by migration can still be likely for patches 100km apart, 
but the shared basis may be hard to detect, as the length of shared haplotype can be quite short.

Furthermore, given the relatively small radius of the patches of dark substrate (often $<10$km),
we should expect the haplotype to be rapidly whittled down within patches after establishment (equation \ref{eqn:rec_rate_within}).
Indeed, if we take the population density to be roughly $\rho \approx 10$ per hectare (M.~Nachman, personal communication),
and the variance in offspring number to be $\xi^2 = 3$ (mostly to get round numbers),
then the shared haplotype length within patches from equation \eqref{eqn:within_haplotype_length}
is only expected to be on the order of hundreds or thousands of bases
($10^{-3}$--$10^{-4}$cM).
This suggests that unless the migration has happened fairly recently \plr{age of the lava flows?}
then it will be difficult to identify shared haplotypes
(how recently depends on the background level of LD in the population).

@@

@@
